Gravitational field calculation in spherical coordinates using variable densities in depth

### 2018-09-21 manuscript/manuscript.tex
Fixed some american to british word spelling

\begin{figure}
\centering
\includegraphics[width=0.6\linewidth]{figures/tesseroid-uieda.pdf}
\caption{
    A tesseroid (spherical prism) in a geocentric spherical coordinate system, with a
    computation point $P$ and its local north oriented Cartesian coordinate system.
    After \citet{Uieda2015}.
}
\label{fig:tesseroid}
\end{figure}

@@

The following algorithm computes the gravitational fields at a given point at $(r, \phi,
\lambda)$ due to a tesseroid with its geometric center at $(r_t, \phi_t, \lambda_t)$ and
dimensions $L_r$, $L_\phi$, and $L_\lambda$ given by

@@

\noindent
in which $D$ is a positive scalar called the distance-size ratio and $d$ is the distance
between the computation point and the geometric center of the tesseroid

@@

\begin{figure*}
\centering
\includegraphics[width=\linewidth]
    {figures/density-based-discretization-algorithm.pdf}
\caption{
    Example application of the density-based discretization algorithm to a non-linear
    density function.
    (a) the normalised density function $\rho_n(r')$ (blue), current boundaries of the
    tesseroid (orange dots), and the linear density function $\rho_l(r')$ (orange line).
    The dashed red line represents the maximum density difference $\Delta \rho (r')$ at
    which the tesseroid would be divided (assuming that the inequality
    \ref{eq:delta-density} is not satisfied).
    (b) second iteration of the algorithm with a new linear density function and maximum
    density difference. The tesseroid would be divided at the depth indicated by the
    dashed red line.
    (c) third iteration of the algorithm.
    (d) final output of the density-based discretization, assuming that all four new
    tesseroids satisfy inequality \ref{eq:delta-density}.
}
\label{fig:density-discretization-algorithm}
\end{figure*}

@@

\begin{figure*}
\centering
\includegraphics[width=\linewidth]{figures/linear-D.pdf}
\caption{
    Differences between the gravitational fields generated by the tesseroid model
    and the analytical solution for a thin (left) and a thick (right) spherical shell.
    Both models have a linear density function of radius (Eq. \ref{eq:density-linear}).
    The computations were performed on the four combinations described in
    Table \ref{tab:grids}.
    Due to the linearity of the density function, the density-based discretization
    algorithm is not applied.
    Differences are reported as a percentage of the shell values.
    The horizontal dashed black line represents a target difference of 0.1\%.
}
\label{fig:D-linear}
\end{figure*}

@@

\begin{figure}
\centering
\includegraphics[width=0.5\linewidth]
    {figures/grid-search.pdf}
\caption{
    Numerical error exploration in the $D$-$\delta$ space.
    The percentage difference values were obtained from the comparison
    between the analytical solution and the numerical approximation of
    the gravitational fields ($V$, $g_z$ and $g_{zz}$) generated by a
    spherical shell with an exponential density function (Eq. \ref{eq:density-exp}).
    These comparisons were carried out on the ``Big Grid'' (Table
    \ref{tab:grids}), with a spherical shell of
    35km of thickness, and density function with $b$ of 1km.
    The points inside the dashed line are the ones that present an
    error lower than 0.1\%.
    }
\label{fig:grid-search}
\end{figure}

@@

\begin{figure*}
\centering
\includegraphics[width=\linewidth]{figures/exponential-delta.pdf}
\caption{
    Numerical error for different exponential density functions and values of the delta
    ratio.
    The top panels show the exponential density functions for different values of $b$
    (Eq. \ref{eq:density-exp}).
    The error computations were performed on the ``Big Grid'' configuration described in
    Table \ref{tab:grids}, with fixed values of the distance-size ratio
    $D$ obtained for the linear density case ($D=1, 2, 8$ for the
    potential, gradient components, and tensor components, respectively).
    The difference is reported as a percentage of the shell values.
    }
\label{fig:delta-exponential}
\end{figure*}

@@

\begin{figure*}
\centering
\includegraphics[width=\linewidth]{figures/neuquen-basin.pdf}
\caption{
    Gravitational effects of the Nequ\'en sedimentary basin modelled
    using tesseroids with an exponential density function of depth.
    (a) Topography of the Neuqu\'en Basin (in km) and its location in South America,
    (b) thickness of the sedimentary basin \citep[in meters;][]{Heine2007},
    (c) the exponential density function used to model the basin
        (depth in km and density in kg/m$^3$),
    (d)-(l) computed gravitational fields: potential $V$ (in J/kg), gradient
    components $g_x$, $g_y$ and $g_z$ (in mGal), and Marussi tensor components
    (in Eötvös), calculated at 50km of height over the ellipsoid.
}
\label{fig:neuquen-basin}
\end{figure*}

@@

Because there is no direct relation between the values assigned to $D$ and $\delta$ and
the error of the computation, we empirically determined the optimum values for these
parameters.
These values minimize the computational load while maintaining the numerical error below
0.1\% of an analytical solution.
The density functions used to establish the optimum values were a linear and an
exponential function.
The linear function represents the smoothest variation of density and does not require
density-based discretization at all.
We analyzed the error for exponential functions ranging from smooth to sharp variations
in depth to test the accuracy of the density-based discretization.
Through this analysis, we have obtained optimal values for the distance-size ratio of
1, 2 and 8 for the potential, its gradient, and the Marussi tensor components,
respectively.
These values are in agreement with previous results for a homogeneous density tesseroid.
The optimum $\delta$ values are 0.2 for the potential and its gradient components, while
a $\delta = 0.01$ must be used for the Marussi tensor components in order to ensure a
0.1\% precision.
A wide range of exponential functions were used to infer these results and they are
conservative estimates, chosen to prioritize accuracy over speed of computations.
Therefore, they can be safely generalized to other common depth-density relations, like
quadratic functions, without loss of accuracy.

@@

\begin{figure}
\centering
\includegraphics[width=0.5\linewidth]{figures/spherical-shell.pdf}
\caption{
    Spherical shell with inner and outer radii $R_1$ and $R_2$, respectively.
    The computation point $Q$ is located on the $z$ axis at a distance $r$ from
    the origin of the coordinates system.
    For our purposes we will assume that $Q$ is outside of the shell,
    i.e. $r > R_2$.
}
\label{fig:spherical-shell}
\end{figure}

>>>>

\begin{figure}
\centreing
\includegraphics[width=0.6\linewidth]{figures/tesseroid-uieda.pdf}
\caption{
    A tesseroid (spherical prism) in a geocentric spherical coordinate system, with a
    computation point $P$ and its local north oriented Cartesian coordinate system.
    After \citet{Uieda2015}.
}
\label{fig:tesseroid}
\end{figure}

@@

The following algorithm computes the gravitational fields at a given point at $(r, \phi,
\lambda)$ due to a tesseroid with its geometric centre at $(r_t, \phi_t, \lambda_t)$ and
dimensions $L_r$, $L_\phi$, and $L_\lambda$ given by

@@

\noindent
in which $D$ is a positive scalar called the distance-size ratio and $d$ is the distance
between the computation point and the geometric centre of the tesseroid

@@

\begin{figure*}
\centreing
\includegraphics[width=\linewidth]
    {figures/density-based-discretization-algorithm.pdf}
\caption{
    Example application of the density-based discretization algorithm to a non-linear
    density function.
    (a) the normalised density function $\rho_n(r')$ (blue), current boundaries of the
    tesseroid (orange dots), and the linear density function $\rho_l(r')$ (orange line).
    The dashed red line represents the maximum density difference $\Delta \rho (r')$ at
    which the tesseroid would be divided (assuming that the inequality
    \ref{eq:delta-density} is not satisfied).
    (b) second iteration of the algorithm with a new linear density function and maximum
    density difference. The tesseroid would be divided at the depth indicated by the
    dashed red line.
    (c) third iteration of the algorithm.
    (d) final output of the density-based discretization, assuming that all four new
    tesseroids satisfy inequality \ref{eq:delta-density}.
}
\label{fig:density-discretization-algorithm}
\end{figure*}

@@

\begin{figure*}
\centreing
\includegraphics[width=\linewidth]{figures/linear-D.pdf}
\caption{
    Differences between the gravitational fields generated by the tesseroid model
    and the analytical solution for a thin (left) and a thick (right) spherical shell.
    Both models have a linear density function of radius (Eq. \ref{eq:density-linear}).
    The computations were performed on the four combinations described in
    Table \ref{tab:grids}.
    Due to the linearity of the density function, the density-based discretization
    algorithm is not applied.
    Differences are reported as a percentage of the shell values.
    The horizontal dashed black line represents a target difference of 0.1\%.
}
\label{fig:D-linear}
\end{figure*}

@@

\begin{figure}
\centreing
\includegraphics[width=0.5\linewidth]
    {figures/grid-search.pdf}
\caption{
    Numerical error exploration in the $D$-$\delta$ space.
    The percentage difference values were obtained from the comparison
    between the analytical solution and the numerical approximation of
    the gravitational fields ($V$, $g_z$ and $g_{zz}$) generated by a
    spherical shell with an exponential density function (Eq. \ref{eq:density-exp}).
    These comparisons were carried out on the ``Big Grid'' (Table
    \ref{tab:grids}), with a spherical shell of
    35km of thickness, and density function with $b$ of 1km.
    The points inside the dashed line are the ones that present an
    error lower than 0.1\%.
    }
\label{fig:grid-search}
\end{figure}

@@

\begin{figure*}
\centreing
\includegraphics[width=\linewidth]{figures/exponential-delta.pdf}
\caption{
    Numerical error for different exponential density functions and values of the delta
    ratio.
    The top panels show the exponential density functions for different values of $b$
    (Eq. \ref{eq:density-exp}).
    The error computations were performed on the ``Big Grid'' configuration described in
    Table \ref{tab:grids}, with fixed values of the distance-size ratio
    $D$ obtained for the linear density case ($D=1, 2, 8$ for the
    potential, gradient components, and tensor components, respectively).
    The difference is reported as a percentage of the shell values.
    }
\label{fig:delta-exponential}
\end{figure*}

@@

\begin{figure*}
\centreing
\includegraphics[width=\linewidth]{figures/neuquen-basin.pdf}
\caption{
    Gravitational effects of the Nequ\'en sedimentary basin modelled
    using tesseroids with an exponential density function of depth.
    (a) Topography of the Neuqu\'en Basin (in km) and its location in South America,
    (b) thickness of the sedimentary basin \citep[in meters;][]{Heine2007},
    (c) the exponential density function used to model the basin
        (depth in km and density in kg/m$^3$),
    (d)-(l) computed gravitational fields: potential $V$ (in J/kg), gradient
    components $g_x$, $g_y$ and $g_z$ (in mGal), and Marussi tensor components
    (in Eötvös), calculated at 50km of height over the ellipsoid.
}
\label{fig:neuquen-basin}
\end{figure*}

@@

Because there is no direct relation between the values assigned to $D$ and $\delta$ and
the error of the computation, we empirically determined the optimum values for these
parameters.
These values minimize the computational load while maintaining the numerical error below
0.1\% of an analytical solution.
The density functions used to establish the optimum values were a linear and an
exponential function.
The linear function represents the smoothest variation of density and does not require
density-based discretization at all.
We analysed the error for exponential functions ranging from smooth to sharp variations
in depth to test the accuracy of the density-based discretization.
Through this analysis, we have obtained optimal values for the distance-size ratio of
1, 2 and 8 for the potential, its gradient, and the Marussi tensor components,
respectively.
These values are in agreement with previous results for a homogeneous density tesseroid.
The optimum $\delta$ values are 0.2 for the potential and its gradient components, while
a $\delta = 0.01$ must be used for the Marussi tensor components in order to ensure a
0.1\% precision.
A wide range of exponential functions were used to infer these results and they are
conservative estimates, chosen to prioritize accuracy over speed of computations.
Therefore, they can be safely generalized to other common depth-density relations, like
quadratic functions, without loss of accuracy.

@@

\begin{figure}
\centreing
\includegraphics[width=0.5\linewidth]{figures/spherical-shell.pdf}
\caption{
    Spherical shell with inner and outer radii $R_1$ and $R_2$, respectively.
    The computation point $Q$ is located on the $z$ axis at a distance $r$ from
    the origin of the coordinates system.
    For our purposes we will assume that $Q$ is outside of the shell,
    i.e. $r > R_2$.
}
\label{fig:spherical-shell}
\end{figure}
